Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-06-26 elsevier_MANUSCRIPT.qmd
new submission docs

To address these limitations, this study focuses on Arizona’s arid and semi-arid basins, where groundwater–surface water interactions remain poorly characterized due to sparse monitoring and regionally distinct hydrologic behavior. We develop a region-specific machine learning model using eXtreme Gradient Boosting (XGBoost) to estimate long-term BFI (1991–2020) across ungauged basins, incorporating basin hydrogeology and hydroclimate predictors. In parallel, we assess trends in BFI at instrumented sites alongside coincident changes in precipitation, temperature, and reference evapotranspiration. Together, these analyses provide new insight into the spatial patterns and temporal dynamics of baseflow in a water-stressed region increasingly affected by climate change. The study objectives are to (1) quantify long-term spatial and temporal trends in BFI across Arizona and (2) develop a predictive model to estimate BFI in ungauged basins based on geospatial and climate variables.

@@

Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors]. XGBoost is a gradient boosting algorithm that builds an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost]. The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins. These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|-------------------------|-------|-------|-------|-------|-------|-------|-------|
| Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |
| linear model | 7724 | 0.248 | 0.046 | 0.214 | 0.173 | 0.247 | \~ 0 |
| Inverse Distance Weighted (LOO CV) | 7724 | 0.449 | 0.029 | 0.170 | 0.131 | 0.448 | 1.86 |
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

>>>>

To address these limitations, this study focuses on Arizona’s arid and semi-arid basins, where groundwater–surface water interactions remain poorly characterized due to sparse monitoring and regionally distinct hydrologic behavior. We develop a region-specific machine learning model using eXtreme Gradient Boosting (XGBoost) to estimate long-term BFI (1991–2020) across ungauged basins, incorporating basin hydrogeology and hydroclimate predictors. To our knowledge, this is the first map of spatially distributed BFI in Arizona. In parallel, we assess trends in BFI at instrumented sites alongside coincident changes in precipitation, temperature, and reference evapotranspiration. Together, these analyses provide new insight into the spatial patterns and temporal dynamics of baseflow in a water-stressed region increasingly affected by climate change. The study objectives are to (1) quantify long-term spatial and temporal trends in BFI across Arizona and (2) develop a predictive model to estimate BFI in ungauged basins based on geospatial and climate variables.

@@

Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors]. XGBoost is a gradient boosting algorithm that builds an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost]. To our knowledge this is the first time that this advanced, decision-tree algorithm has been used in hydrologic prediction. The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins. These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |
| linear model | 7724 | 0.248 | 0.046 | 0.214 | 0.173 | 0.247 | \~ 0 |
| Inverse Distance Weighted (LOO CV) | 7724 | 0.449 | 0.029 | 0.170 | 0.131 | 0.448 | 1.86 |
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |
